{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}

### 2017-10-20 sftrends_manu.Rnw
rafi edits done

Long-term monitoring data are powerful sources of information that can facilitate descriptions of water quality change in coastal regions. A comprehensive water quality monitoring program has been in place in the Delta for several decades \citep[\cref{fig:delt_map},][]{IEP13}.  Although these data have been used extensively, water quality trends covering the full spatial and temporal coverage of the monitoring dataset have not been systematically evaluated. Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.  An evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that the majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported \citep{Novick15}. Seasonal and annual changes in the delivery of water inflows and water exports directly from the system can also obscure trends \citep{Jassby00,Jassby08}.

@@

Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten active sampling stations in the Delta \citep[\cref{fig:delt_map,tab:stdescrp},][]{IEP13}.  Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations cover all of the major inflows and outflows to the Delta and were selected for analysis based on the continuity of the period of observation \citep{Jabusch09}. Although many other stations are available for the region, the stations were chosen because they are actively maintained by the regional monitoring program and they capture dominant seasonal and annual modes of nitrogen variability characteristic of the region \citep{Jabusch16}. Time series were complete for all stations except for an approximate ten year gap from 1996-2004 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).

Daily flow estimates for the Delta region were obtained from the Dayflow software program \citep{IEP16}. The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow. Daily flow estimates were matched with the corresponding sample dates for the nutrient data. Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation, given the stronger tidal influence at these stations.  Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \citep{Beck15}.  Salinity was not used as a covariate at the interior and peripheral stations becuase initial analyses showed improved performance using flow estimates from the Sacramento and San Joaquin rivers.

@@

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual ($\beta_1$) or seasonal trend ($\beta_3$, $\beta_4$), and $Q$ is the flow variable (either flow or salinity depending on station).  Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters (i.e., $\beta_0$, \ldots, $\beta_4$) at each observation point in the time series ($n$ ranging from \Sexpr{minobs} at \Sexpr{minsta} to \Sexpr{maxobs} at \Sexpr{maxsta}).  Rather than fitting a global model to the entire time series, one regression is fit to every observation.  Observations within a window for each regression are weighted relative to annual, seasonal, and flow distances from the observation at the center of the window.  Observations with distances farther from the center (i.e., greater time and different flow values from the center) have less weight during parameter estimation for each regression.  This approach allows for a type of smoothing where the observed fit is specific to the data characteristics within windows. Models applied herein were based on a tidal adaptation of the original method that can use either flow or salinity estimates as nutrient predictors \citep{Beck15}.  All models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \citep{Tobin58}.  Model predictions were evaluated as monthly values or as annual values that averaged monthly results within each water year (October to September). All analyses used the WRTDStidal package for the R statistical programming language \citep{Beck17b,RDCT17}.  The default model fitting procedures were used that set half-window widths as six months for seasonal weights, ten years for annual weights, and half the range of salinity or flow in the input data for $Q$ weights.

@@

@@

@
A comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends. For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period (annual or seasonal aggregations), \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant. This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter. Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that \Sexpr{eng(sum(chgtb[c('morepos', 'moreneg')]))} comparisons showed an increase in magnitude (more negative or more positive) and \Sexpr{eng(sum(chgtb[c('lesspos', 'lessneg')]))} had a decrease (less positive or less negative) compared to observed trends.  \Sexpr{cap(eng(chgtb['postoneg']))} comparisons showed a trend reversal from positive to negative estimated change, \Sexpr{eng(chgtb['zerotoneg'])} sites went from no change to negative estimated change, and \Sexpr{eng(chgtb['zerotopos'])} site went from no change to a positive trend for the flow-normalized results. Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}). The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.

@@

Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources  \citep{Cornwell14,Novick15}.  Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate. The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \citep{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream. A modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
\item a shift in the ratio of the components of \ac{din} from the \ac{wwtp} before/after upgrade, and
\item a flow-normalized annual trend at P8 to show a change concurrent with \ac{wwtp} upgrades.
\end{inparaenum}

@@

Effluent measured from 2003 to 2009 from the Stockton \ac{wwtp} had a gradual reduction in ammonium concentration relative to total \ac{din} (\cref{fig:stock}).  Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite. As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.  Interestingly, nitrite/nitrate concentrations at P8 also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade (\cref{fig:stock}).  Percent changes from seasonal Kendall tests on flow-normalized results (\cref{tab:p8chg}) showed that both nitrogen species increased prior to \ac{wwtp} upgrades (\Sexpr{scinot(no_bf[1, 'perchg'], 1, 1)}\% per year for nitrite/nitrate, \Sexpr{scinot(nh_bf[1, 'perchg'], 1, 1)}\% for ammonium), followed by decreases after upgrades (\Sexpr{scinot(no_af[1, 'perchg'], 1, 1)}\% for nitrite/nitrate, \Sexpr{scinot(nh_af[1, 'perchg'], 1, 1)}\% for ammonium).  Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate (\Sexpr{scinot(no_bf[3, 'perchg'], 1, 1)}\%) and in the fall for ammonium (\Sexpr{scinot(nh_bf[4, 'perchg'], 1, 1)}\%). Similarly, seasonal reductions post-upgrade were largest in the summer for nitrite/nitrate (\Sexpr{scinot(no_af[3, 'perchg'], 1, 1)}\%) and largest for ammonium in the winter (\Sexpr{scinot(nh_af[5, 'perchg'], 1, 1)}\%).

Relationships of nitrite/nitrate with flow described by \ac{wrtds} showed an inverse flow and concentration dynamic with flushing or dilution at higher flow (\cref{fig:p8trnds}b).  Seasonal variation was even more apparent for ammonium, although both nitrite/nitrate and ammonium typically had the highest concentrations at low flow in the winter (January).  Additionally, strength of the flow/nutrient relationship changed between years. Nitrite/nitrate typically had the strongest relationship with flow later in the time series (i.e., larger negative slope), whereas ammonium had the strongest relationship with flow around 2000 in January.

@@

The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}i), with lower \ac{chla} associated with higher biomass, confirming results from earlier studies \citep{Alpine92,Thompson08}.  However, the effect of flow on both clams and phytoplankton as a top-down or bottom-up control changed throughout the time series.  The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}j), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series. Following clam invasion, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow. The increase in clam abundance was concurrent with decline in \ac{chla} concentration, although variation in abundance between years was also observed.  Clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\ref{fig:clmchl}a). In the same years, \ac{wrtds} predictions for \ac{chla} were higher than the flow-normalized component (\cref{fig:clmchl}c), which further suggests a link between increased flow and phytoplankton production.

@@

% obs v flow norm trends
Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions.  Our results demonstrated the potential to interpret different trends if flow effects are not considered, where the difference could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}). A more comprehensive evaluation of flow in the Delta demonstrated that flow contributions of different end members vary considerably over time at each station \citep{Novick15}.  For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.   For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.  Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location. Output from the Dayflow software program \citep{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.

@@

A general conclusion from the analysis of nitrogen trends at P8 is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January.  These dynamics are difficult to characterize from the observed time series, and further, results from \ac{wrtds} can be used to develop additional hypotheses of factors that influence nutrient concentrations at P8. For example, estimated ammonium concentrations in July were low for all flow levels, which suggests either nitrogen inputs were low in the summer or nitrogen was available but uptake by primary consumers and bacterial processing were high. Seasonal patterns in the relationship between flow and nitrite/nitrate were not as dramatic as compared to ammonium, and in particular, low-flow events in July were associated with higher concentrations.  This could suggest that ammonium concentrations at P8 are driving phytoplankton production at low flow during warmer months, and not nitrite/nitrate given the higher estimated concentrations in July at low flow. As such, these simple observations provide quantitative support of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.  Additional research could investigathese hypotheses to better describe mechanisms of change as a basis for more informed management.

In addition to the above comments, our results for Suisun Bay provide additional descriptions of change in production as it relates to flow, grazing, and nitrogen ratios.  In general, our
results suggested that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.  This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.  Anontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P. amurensis}.  Flow effects on phytoplankton production have also changed over time. In the absence of benthic grazing prior to invasion, \ac{chla} production was limited at low flow as less nutrients were exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the larger bay (\cref{fig:clmchl}j). Recent years have shown a decrease in overall chlorophyll, with particularly low concentrations at low flow (high salinity).  As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure. These relationships have been suggested by others \citep{Cloern83,Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.

@@

As demonstrated by both case studies and the overall trends across all stations, water quality dynamics in the Delta are complex and driven by multiple factors that change through space and time.  At a minimum, \ac{wrtds} provides a description of change by focusing on high-level forcing factors that explicitly account for annual, seasonal, and flow effects on trend interpretations.  We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trends have more power to quantify change.  Moreover, trends in nutrient loads from point sources in the Delta have previously been described, e.g., Sacramento \ac{wwtp} increases \citep{Jassby08} and exports to Suisun Bay \citep{Novick14}. The results from \ac{wrtds} demonstrating these changes are not unexpected, and consequently, we are not detracting from the potential implications of such increases. The important conclusion is that the physical/hydrological and biogeochemical factors that influence nutrient cycling and ambient concentrations in the Bay-Delta, and changes to those factors, are substantial enough that they can be comparable in magnitude to anthropogenic load increases or comparable to the effects of management actions to decrease nutrient levels. Therefore, methods that adjust for the effects of these factors are critical when studying long-term records to assess the impacts or effectiveness of load increases or management actions, respectively.

Combined with additional data, \ac{wrtds} results can support hypotheses that lead to a more comprehensive understanding of ecosystem dynamics. Additional factors to consider include the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.  For example, a more rigorous matching of flow time series with water quality observations at each station that considers varying source contributions over time could provide a more robust description of flow-normalized results.  Alternative methods for time series analysis could also be used to address a wider range of questions, particularly those with more generic structural forms that can explicitly include additional variables (e.g., generalized additive models, \cite{Beck17}).  Overall, statistical interpretations of multiple factors can provide a basis for quantitative links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.

@@

cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(5, 7)] %>% 
  rev
ylab <- expression(paste('DIN (mg ', L^-1, ')'))

@@

p1 <- ggplot(toplo, aes(x = Date)) +
  geom_area(aes(y = val, group = var, fill = var), stat = 'identity', position = 'stack', alpha = 0.75) + 
  scale_x_date(expand = c(0, 0)) + 
  scale_y_continuous(ylab, expand = c(0, 0)) + 
  facet_grid(Site_Code~., scales = 'free_y') +
  scale_fill_manual(values = cols, labels = leglabs) +
  mythm %+replace% 
  theme(axis.text.x = element_blank())

>>>>

Long-term monitoring data are powerful sources of information that can facilitate descriptions of water quality change in coastal regions. A comprehensive water quality monitoring program has been in place in the Delta for several decades \citep[\cref{fig:delt_map},][]{IEP13}.  Although these data have been used extensively, water quality trends covering the full spatial and temporal coverage of the monitoring dataset have not been systematically evaluated. Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.  An evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that the majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported to the estuary \citep{Novick15}. Seasonal and annual changes in the delivery of water inflows and water exports directly from the system can also obscure trends \citep{Jassby00,Jassby08}.

@@

Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten active sampling stations in the Delta \citep[\cref{fig:delt_map,tab:stdescrp},][]{IEP13}.  Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations cover all of the major inflows and outflows to the Delta and were selected for analysis based on the continuity of the period of observation \citep{Jabusch09}. Although many other stations are available for the region, the stations were chosen because they are actively maintained by the regional monitoring program and they capture dominant seasonal and annual modes of nitrogen variability characteristic of the region \citep{Jabusch16}. Time series were complete for all stations except for an approximate ten year gap from 1996-2004 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were below the deteciton limit (left-censored), although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).

\ac{wrtds} models require flow data paired with nutrient data. At the peripheral and interior stations, daily flow estimates were matched with the corresponding sample dates for the nutrient data. Daily flow estimates were obtained from the Dayflow software program \citep{IEP16}.  The Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series. Salinity was used as a proxy for flow at sites D4, D6, and D7 where tidal influences were much stronger. Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \citep{Beck15}.  Models were evaluated using salinity at the interior and peripheral stations but performance was reduced relative to models that used daily flow estimates.

@@

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual ($\beta_1$) or seasonal trend ($\beta_3$, $\beta_4$), and $Q$ is the flow variable (either flow or salinity depending on station).  Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters (i.e., $\beta_0$, \ldots, $\beta_4$) at each observation point in the time series ($n$ ranging from \Sexpr{minobs} at \Sexpr{minsta} to \Sexpr{maxobs} at \Sexpr{maxsta}).  Rather than fitting a global model to the entire time series, one regression is fit to every observation.  Observations within a window for each regression are weighted relative to annual, seasonal, and flow distances from the observation at the center of the window.  Observations with distances farther from the center (i.e., greater time and different flow values from the center) have less weight during parameter estimation for each regression.  This approach allows for a type of smoothing where the observed fit is specific to the data characteristics within windows. Models applied herein were based on a tidal adaptation of the original method that can use either flow or salinity estimates as nutrient predictors \citep{Beck15}.  All models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data below the detection limit \citep{Tobin58}.  Model predictions were evaluated as monthly values or as annual values that averaged monthly results within each water year (October to September). All analyses used the WRTDStidal package for the R statistical programming language \citep{Beck17b,RDCT17}.  The default model fitting procedures were used that set half-window widths as six months for seasonal weights, ten years for annual weights, and half the range of salinity or flow in the input data for $Q$ weights.

@@

Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources  \citep{Cornwell14,Novick15}.  Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate. The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 \citep{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream. An increase and decrease of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
\item a shift in the ratio of the components of \ac{din} from the \ac{wwtp} before/after upgrade, and
\item a flow-normalized annual trend at P8 to show a change concurrent with \ac{wwtp} upgrades.
\end{inparaenum}

@@

@
A comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends. For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period (annual or seasonal aggregations), \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant ($p > 0.05$) with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant. This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter. Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that \Sexpr{eng(sum(chgtb[c('morepos', 'moreneg')]))} comparisons showed an increase in magnitude (more negative or more positive) and \Sexpr{eng(sum(chgtb[c('lesspos', 'lessneg')]))} had a decrease (less positive or less negative) compared to observed trends.  \Sexpr{cap(eng(chgtb['postoneg']))} comparisons showed a trend reversal from positive to negative estimated change, \Sexpr{eng(chgtb['zerotoneg'])} sites went from no change to negative estimated change, and \Sexpr{eng(chgtb['zerotopos'])} site went from no change to a positive trend for the flow-normalized results. Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}). The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.

@@

@@

Effluent measured from 2003 to 2009 from the Stockton \ac{wwtp} had similar \ac{din} concentrations before and after upgrades, whereas ammonium concentrations were greatly reduced (\cref{fig:stock}).  Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite. As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.  Nitrite/nitrate concentrations at P8 also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade (\cref{fig:stock}).  Percent changes from seasonal Kendall tests on flow-normalized results (\cref{tab:p8chg}) showed that both nitrogen species increased prior to \ac{wwtp} upgrades (\Sexpr{scinot(no_bf[1, 'perchg'], 1, 1)}\% per year for nitrite/nitrate, \Sexpr{scinot(nh_bf[1, 'perchg'], 1, 1)}\% for ammonium), followed by decreases after upgrades (\Sexpr{scinot(no_af[1, 'perchg'], 1, 1)}\% for nitrite/nitrate, \Sexpr{scinot(nh_af[1, 'perchg'], 1, 1)}\% for ammonium).  Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate (\Sexpr{scinot(no_bf[3, 'perchg'], 1, 1)}\%) and in the fall for ammonium (\Sexpr{scinot(nh_bf[4, 'perchg'], 1, 1)}\%). Similarly, seasonal reductions post-upgrade were largest in the summer for nitrite/nitrate (\Sexpr{scinot(no_af[3, 'perchg'], 1, 1)}\%) and largest for ammonium in the winter (\Sexpr{scinot(nh_af[5, 'perchg'], 1, 1)}\%).

Nitrogen concentrations varied with flow, although relationships depended on season and year. Relationships of nitrite/nitrate with flow described by \ac{wrtds} showed flushing or dilution at higher flow (\cref{fig:p8trnds}b).  Seasonal variation was even more apparent for ammonium, although both nitrite/nitrate and ammonium typically had the highest concentrations at low flow in the winter (January).  Additionally, strength of the flow/nutrient relationship changed between years. Nitrite/nitrate typically had the strongest relationship with flow later in the time series (i.e., larger negative slope), whereas ammonium had the strongest relationship with flow around 2000 in January.

@@

The relationship of \ac{chla} with clam biomass was significant (\cref{fig:clmchl}i), with lower \ac{chla} associated with higher biomass. However, the effect of flow on both clams and phytoplankton as a top-down or bottom-up control changed throughout the time series.  The \ac{chla}/flow relationship showed that increasing flow (decreasing salinity) was associated with a slight increase in \ac{chla} followed by a decrease early in the time series (\cref{fig:clmchl}j), whereas overall \ac{chla} was lower but a positive association with flow (negative with salinity) was observed later in the time series. Following clam invasion, \ac{chla} concentrations were reduced by grazing but showed a positive and monotonic relationship with increasing flow. The increase in clam abundance was concurrent with decline in \ac{chla} concentration, although variation in abundance between years was also observed.  Clam abundance was reduced during high flow years in the late 1990s, 2006, and 2011 (\ref{fig:clmchl}a). In the same years, \ac{wrtds} predictions for \ac{chla} were higher than the flow-normalized component (\cref{fig:clmchl}c), which further suggests a link between increased flow and phytoplankton production.

@@

% obs v flow norm trends
Our results demonstrated the potential to interpret different trends if flow effects are not considered, where the difference could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}). Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions. An alternative evaluation of flow in the Delta demonstrated that flow contributions from different sources vary considerably over time at each station \citep{Novick15}.  For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.   For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.  Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location. Output from the Dayflow software program \citep{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.

@@

Seasonal timings of water quality improvements and the link to flow changes are difficult to characterize from the observed time series. \ac{wrtds} models were used to characterize these changes at P8 and to develop additional hypotheses of factors that influence nutrient concentrations. A general conclusion is that ammonium reductions were concurrent with \ac{wwtp} upgrades, but the reduction was most apparent at low-flow in January. Estimated ammonium concentrations in July were low for all flow levels, which suggests either nitrogen inputs were low in the summer or nitrogen was available but uptake by primary consumers and bacterial processing were high. Seasonal patterns in the relationship between flow and nitrite/nitrate were not as dramatic as compared to ammonium, and in particular, low-flow events in July were associated with higher concentrations.  This could suggest that ammonium concentrations at P8 are driving phytoplankton production at low flow during warmer months, and not nitrite/nitrate given the higher estimated concentrations in July at low flow. As such, these simple observations provide quantitative support of cause and effect mechanisms of nutrient impacts on potentially adverse environmental conditions as they relate to nutrient-related source controls upstream.  Additional research could investigate these hypotheses to better describe mechanisms of change as a basis for more informed management.

The results for Suisun Bay provide additional descriptions of change in production as it relates to flow, grazing, and nitrogen ratios.  In general, clam biomass was associated with a decrease in \ac{chl} concentration, as shown by others \citep{Alpin92,Thompson08}. Our results also suggested that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.  This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.  Anontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P. amurensis}.  Flow effects on phytoplankton production have also changed over time. In the absence of benthic grazing prior to invasion, \ac{chla} production was limited at low flow as less nutrients were exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the estuary (\cref{fig:clmchl}j). Recent years have shown a decrease in overall chlorophyll, with particularly low concentrations at low flow (high salinity).  As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure. These relationships have been suggested by others \citep{Cloern83,Alpine92,Parchaso02,Jassby08}, although the precise mechanism demonstrated by \ac{wrtds} provides a quantitative description of factors that drive water quality in the Delta.

@@

As demonstrated by both case studies and the overall trends across all stations, water quality dynamics in the Delta are complex and driven by multiple factors that change through space and time. \ac{wrtds} models can facilitate descriptions of change by focusing on high-level forcing factors that explicitly account for annual, seasonal, and flow effects on trend interpretations.  We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trends have more power to quantify change.  The results from \ac{wrtds} are also consistent with described trends of nutrient loads from point sources (e.g., Sacramento \ac{wwtp} increases \citep{Jassby08} and exports to Suisun Bay \citep{Novick14}) demonstrating that these changes are not unexpected.  Consequently, we are not detracting from the potential implications of such increases. The important conclusion is that the physical/hydrological and biogeochemical factors that influence nutrient cycling and ambient concentrations in the Bay-Delta, and changes to those factors, are substantial enough that they can be comparable in magnitude to anthropogenic load increases or comparable to the effects of management actions to decrease nutrient levels. Therefore, methods that adjust for the effects of these factors are critical when studying long-term records to assess the impacts or effectiveness of load increases or management actions, respectively.

Combined with additional data, our results can support hypotheses that lead to a more comprehensive understanding of ecosystem dynamics. Additional factors to consider include the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.  For example, a more rigorous matching of flow time series with water quality observations at each station that considers varying source contributions over time could provide a more robust description of flow effects.  Alternative methods could also be used to address a wider range of questions, particularly those with more generic structural forms that can explicitly include additional variables (e.g., generalized additive models, \cite{Beck17}).  Overall, quantitative interpretations of multiple factors can provide a basis for identifying links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.

@@

cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(5, 7)] %>% 
  rev
ylab <- expression(paste('log - DIN (mg ', L^-1, ')'))

@@

p1 <- ggplot(toplo, aes(x = Date)) +
  geom_area(aes(y = log10(1 + val), group = var, fill = var), stat = 'identity', position = 'stack', alpha = 0.75) + 
  scale_x_date(expand = c(0, 0)) + 
  scale_y_continuous(ylab, expand = c(0, 0)) + 
  facet_grid(Site_Code~.)+ #, scales = 'free_y') +
  scale_fill_manual(values = cols, labels = leglabs) +
  mythm %+replace% 
  theme(axis.text.x = element_blank())
